---
title: "cBioPortal INDEL Processing Pipeline"
author: "Michael Dunnet"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
```

# 1. Dataset Description

Datasets were downloaded from cBioPortal:

- Colorectal Adenocarcinoma (DFCI, Cell Reports 2016)
- Colorectal Adenocarcinoma (Genentech, Nature 2012)
- TCGA PanCancer Atlas
- Triplets MSK (Genome Biology 2014)
- CAS Shanghai (Cancer Cell 2020)
- HTAN MSK (Nature 2025)
- MSK (Cancer Discovery 2022)
- MSK (Gastroenterology 2020)
- MSK (JCO Precision Oncology 2022)
- ChangKang Project SYSUCC (Nature Communications 2022)
- Rectal Cancer MSK (Nature Medicine 2019)

# 2. Working Directory
Ensure that the correct working directory is set
```{r}
setwd("~/Documents/GitHub/R-Projects/INDEL_Targets_ctRCA/")
```

# 3. Load Data
Make sure the right file name is present. Below is an example
```{r}
APC.dat    <- read.delim("table.tsv")
TP53.dat   <- read.delim("table (1).tsv")
RNF43.dat  <- read.delim("table (2).tsv")
ACVR2A.dat <- read.delim("table (3).tsv")
```

# 4. Keep Relevant Columns
If these columns are not present, update the cols_keep variable to match the correct column indices.
The column indices correspond to:
- Gene
- Protein.Change
- Annotation
- Mutation.Type
- Variant.Type
- Chromosome
- Start.Pos
- End.Pos
- Ref
- Var
- HGVSg
- HGVSc
```{r}
cols_keep <- c(1, 6, 7, 11, 12, 17, 18, 19, 20, 21, 22, 23)

APC.dat    <- APC.dat[, cols_keep]
TP53.dat   <- TP53.dat[, cols_keep]
RNF43.dat  <- RNF43.dat[, cols_keep]
ACVR2A.dat <- ACVR2A.dat[, cols_keep]
```

# 5. Remove Duplicates

```{r}
APC.dat    <- unique(APC.dat)
TP53.dat   <- unique(TP53.dat)
RNF43.dat  <- unique(RNF43.dat)
ACVR2A.dat <- unique(ACVR2A.dat)
```

# 6. Load BED File
Make sure the bed file uses the same genome build as the data from c.BioPortal. Usually this is hg19. If you have a bed file in a different genome build you can use
the USCS genome browser liftover to change between versions. 
```{r}
bed <- read.delim("hglft_genome_3a98b_ffa740.bed", header = FALSE)
colnames(bed) <- c("chr", "start", "end", "Gene")
```

# 7. Filter by Target Regions

```{r}
filter_bed <- function(df) {
  keep <- mapply(function(g, pos) {
    any(bed$Gene == g & pos >= bed$start & pos <= bed$end)
  }, df$Gene, df$Start.Pos)
  
  df[keep, ]
}

APC.dat    <- filter_bed(APC.dat)
TP53.dat   <- filter_bed(TP53.dat)
RNF43.dat  <- filter_bed(RNF43.dat)
ACVR2A.dat <- filter_bed(ACVR2A.dat)
```

# 8. Split INDEL Types

```{r}
split_indels <- function(df) {
  list(
    del = df[df$Variant.Type == "DEL", ],
    ins = df[df$Variant.Type == "INS", ]
  )
}

APC.split    <- split_indels(APC.dat)
TP53.split   <- split_indels(TP53.dat)
RNF43.split  <- split_indels(RNF43.dat)
ACVR2A.split <- split_indels(ACVR2A.dat)
```

# 9. Combine Genes

```{r}
deletion.dat <- bind_rows(
  ACVR2A.split$del,
  APC.split$del,
  TP53.split$del,
  RNF43.split$del
)

insertion.dat <- bind_rows(
  ACVR2A.split$ins,
  APC.split$ins,
  TP53.split$ins,
  RNF43.split$ins
)
```

# 10. Reformat Output

```{r}
reformat <- function(df) {
  df %>%
    rename(
      GENE_NAME = Gene,
      `Mutation CDS` = HGVSc,
      `Mutation AA` = Protein.Change,
      GENOMIC_WT_ALLELE_SEQ = Ref,
      GENOMIC_MUT_ALLELE_SEQ = Var,
      `Mutation Description AA` = Mutation.Type,
      Chrom = Chromosome
    ) %>%
    select(
      GENE_NAME,
      `Mutation CDS`,
      `Mutation AA`,
      GENOMIC_WT_ALLELE_SEQ,
      GENOMIC_MUT_ALLELE_SEQ,
      `Mutation Description AA`,
      Chrom,
      Start.Pos,
      End.Pos
    )
}

deletion.dat.reformat  <- reformat(deletion.dat)
insertion.dat.reformat <- reformat(insertion.dat)
```

# 11. Split Transcript Field

```{r}
deletion.dat.reformat$Transcript <- sub(":.*", "", deletion.dat.reformat$`Mutation CDS`)
deletion.dat.reformat$`Mutation CDS` <- sub(".*:", "", deletion.dat.reformat$`Mutation CDS`)

insertion.dat.reformat$Transcript <- sub(":.*", "", insertion.dat.reformat$`Mutation CDS`)
insertion.dat.reformat$`Mutation CDS` <- sub(".*:", "", insertion.dat.reformat$`Mutation CDS`)
```

# 12. Export Results
After exporting, you may need to re-format the start and end positions to the correct genome version. 
```{r}
write.csv(deletion.dat.reformat,  "Deletion_variants.csv",  row.names = FALSE)
write.csv(insertion.dat.reformat, "Insertion_variants.csv", row.names = FALSE)
```